Study guide: Algorithms and implementations for exponential decay models

### 2015-10-10 doc/src/chapters/alg/slides_alg.tex

% LaTeX Beamer file automatically generated from DocOnce
% https://github.com/hplgit/doconce

%-------------------- begin beamer-specific preamble ----------------------

\documentclass{beamer}

\usetheme{red_shadow}
\usecolortheme{default}

% turn off the almost invisible, yet disturbing, navigation symbols:
\setbeamertemplate{navigation symbols}{}

% Examples on customization:
%\usecolortheme[named=RawSienna]{structure}
%\usetheme[height=7mm]{Rochester}
%\setbeamerfont{frametitle}{family=\rmfamily,shape=\itshape}
%\setbeamertemplate{items}[ball]
%\setbeamertemplate{blocks}[rounded][shadow=true]
%\useoutertheme{infolines}
%
%\usefonttheme{}
%\useinntertheme{}
%
%\setbeameroption{show notes}
%\setbeameroption{show notes on second screen=right}

% fine for B/W printing:
%\usecolortheme{seahorse}

\usepackage{pgf}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{relsize}

\usepackage{fancybox}  % make sure fancybox is loaded before fancyvrb

\usepackage{fancyvrb}
\usepackage{minted} % requires pygments and latex -shell-escape filename
%\usepackage{anslistings}
%\usepackage{listingsutf8}

\usepackage{amsmath,amssymb,bm}
%\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{colortbl}
\usepackage[english]{babel}
\usepackage{tikz}
\usepackage{framed}
% Use some nice templates
\beamertemplatetransparentcovereddynamic

% --- begin table of contents based on sections ---
% Delete this, if you do not want the table of contents to pop up at
% the beginning of each section:
% (Only section headings can enter the table of contents in Beamer
% slides generated from DocOnce source, while subsections are used
% for the title in ordinary slides.)
\AtBeginSection[]
{
  \begin{frame}<beamer>[plain]
  \frametitle{}
  %\frametitle{Outline}
  \tableofcontents[currentsection]
  \end{frame}
}
% --- end table of contents based on sections ---

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\newcommand{\shortinlinecomment}[3]{\note{\textbf{#1}: #2}}
\newcommand{\longinlinecomment}[3]{\shortinlinecomment{#1}{#2}{#3}}

\definecolor{linkcolor}{rgb}{0,0,0.4}
\hypersetup{
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3
    }
\setlength{\parskip}{7pt}  % {1em}

\newenvironment{doconceexercise}{}{}
\newcounter{doconceexercisecounter}
\newenvironment{doconce:movie}{}{}
\newcounter{doconce:movie:counter}

\newcommand{\subex}[1]{\noindent\textbf{#1}}  % for subexercises: a), b), etc

%-------------------- end beamer-specific preamble ----------------------

% Add user's preamble

@@

\raggedbottom
\makeindex

@@

\title{Study guide: Algorithms and implementations for exponential decay models}

@@

\author{Hans Petter Langtangen\inst{1,2}}
\institute{Center for Biomedical Computing, Simula Research Laboratory\inst{1}
\and
Department of Informatics, University of Oslo\inst{2}}
% ----------------- end author(s) -------------------------

\date{Oct 10, 2015
% <optional titlepage figure>
% <optional copyright>
}

\begin{frame}[plain,fragile]
\titlepage
\end{frame}

\section{INF5620 in a nutshell}

\begin{frame}[plain,fragile]
\frametitle{INF5620 in a nutshell}

\label{5620:about}

@@

\noindent
\begin{block}{After the course }
You see a PDE and can't wait to program a method
and visualize a solution! Somebody asks if the solution is right
and you can give a convincing answer.
\end{block}

\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The new official six-point course description}

@@

\begin{itemize}
\pause
 \item can derive methods and implement them to solve frequently
   arising partial differential equations (PDEs) from physics and mechanics.

\pause
 \item have a good understanding of finite difference and finite element
   methods and how they are applied in linear and nonlinear PDE problems.

\pause
 \item can identify numerical artifacts and perform mathematical analysis
   to understand and cure non-physical effects.

\pause
 \item can apply sophisticated programming techniques in Python, combined
   with Cython, C, C++, and Fortran code, to create modern,
   flexible simulation programs.

\pause
 \item can construct verification tests and automate them.

\pause
 \item have experience with project hosting sites (GitHub),
   version control systems (Git), report writing ({\LaTeX}),
   and Python scripting for performing reproducible computational science.
\end{itemize}

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{More specific contents: finite difference methods}

\begin{block}{}
\begin{itemize}
   \item ODEs

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{More specific contents: finite element methods}

\begin{block}{}
\begin{itemize}
   \item stationary diffusion equations $u_{xx}=f$ in 1D

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{More specific contents: nonlinear and advanced problems}

\begin{block}{}
\begin{itemize}
 \item Nonlinear PDEs
\begin{itemize}

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Philosophy: simplify, understand, generalize}

@@

\noindent
This is the power of applied mathematics!
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The exam}

\begin{itemize}
\pause
 \item Oral exam

\pause
 \item 6 problems (topics) are announced two weeks before the exam

\pause
 \item Work out a 20 min presentations (talks) for each problem

\pause
 \item At the exam: throw a die to pick your problem to be presented

\pause
 \item Aids: plots, computer programs

\pause
 \item Why? Very effective way of learning

\pause
 \item Sure? Excellent results over 15 years

\pause
 \item When? Late december
\end{itemize}

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Required software}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Assumed/ideal background}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Start-up example for the course}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Start-up example}

\begin{block}{ODE problem }
\[ u'=-au,\quad u(0)=I,\ t\in (0,T],\]
where $a>0$ is a constant.
\end{block}

Everything we do is motivated by what we need as building blocks for
solving PDEs!
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{What to learn in the start-up example; standard topics}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{What to learn in the start-up example; programming topics}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{What to learn in the start-up example; mathematical analysis}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{What to learn in the start-up example; generalizations}

@@

\noindent
\end{frame}

\section{Finite difference methods}

\begin{frame}[plain,fragile]
\frametitle{Finite difference methods}

\label{decay:fdm}

@@

\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Topics in the first intro to the finite difference method}

\begin{block}{Contents }
\begin{itemize}
 \item How to think about finite difference discretization

@@

\noindent
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{A basic model for exponential decay}

\label{decay:model}

@@

\begin{block}{Observation }
We can learn a lot about numerical methods, computer implementation,
program testing, and real applications of these tools by using
this very simple ODE as example. The teaching principle is to keep the math as
simple as possible while learning computer tools.
\end{block}

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/FE1.png}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The ODE model has a range of applications in many fields}

@@

\noindent
See the \href{{http://tinyurl.com/nclmcng/pub/sphinx-decay/._main_decay008.html}}{text} for details.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The ODE problem has a continuous and discrete version}

\noindent\textbf{Continuous problem.}
\begin{equation}
u' = -au,\ t\in (0,T], \quad u(0)=I
\label{decay:problem}
\end{equation}
(varies with a continuous $t$)

\noindent\textbf{Discrete problem.}
Numerical methods applied to the continuous problem turns it into
a discrete problem

\begin{equation}
u^{n+1} = \mbox{const} u^n, \quad n=0,1,\ldots N_t-1, \quad u^n=I
\label{decay:problem:discrete}
\end{equation}
(varies with discrete mesh points $t_n$)
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The steps in the finite difference method}

\label{decay:schemes:keysteps}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Step 1: Discretizing the domain}

@@

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/fdm_u_ue.png}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{What about a mesh function between the mesh points?}

@@

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/fdm_u_uei.png}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Step 2: Fulfilling the equation at discrete time points}

@@

\index{finite differences}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Step 3: Replacing derivatives by finite differences}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_forward.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Step 3: Replacing derivatives by finite differences}

@@

(Known as \emph{discrete equation}, or \emph{discrete problem},
or \emph{finite difference method/scheme})
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Step 4: Formulating a recursive algorithm}

@@

\begin{block}{The Forward Euler scheme }
Solve wrt $u^{n+1}$ to get the computational formula:

\begin{equation}
u^{n+1} = u^n - a(t_{n+1} -t_n)u^n
\label{decay:FE}
\end{equation}
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Let us apply the scheme by hand}

@@

Ooops - we can find the numerical solution by hand (in this simple
example)! No need for a computer (yet)...
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{A backward difference}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_backward.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Backward Euler scheme}

@@

\begin{equation}
u^{n+1} = \frac{1}{1+ a(t_{n+1}-t_n)} u^n
\label{decay:BE}
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{A centered difference}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_centered_CN.pdf}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Crank-Nicolson scheme; ideas}

\label{decay:schemes:CN}

@@

\[ u(t_{n+\half}) \approx \half(u^n + u^{n+1}) \]
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Crank-Nicolson scheme; result}

@@

\begin{equation}
u^{n+1} = \frac{1-\half a(t_{n+1}-t_n)}{1 + \half a(t_{n+1}-t_n)}u^n
\label{decay:CN}
\end{equation}
This is a Crank-Nicolson (CN) scheme or a midpoint or centered scheme.
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The unifying $\theta$-rule}

\label{decay:schemes:theta}

@@

\begin{equation}
u^{n+1} = \frac{1 - (1-\theta) a(t_{n+1}-t_n)}{1 + \theta a(t_{n+1}-t_n)}
\label{decay:th}
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Constant time step}

@@

\begin{block}{Summary of schemes for constant time step }
\begin{alignat}{2}
u^{n+1} &= (1 - a\Delta t )u^n  & \hbox{Forward Euler}
\label{decay:FE:u}\\ 
u^{n+1} &= \frac{1}{1+ a\Delta t} u^n  & \hbox{Backward Euler}
\label{decay:BE:u}\\ 
u^{n+1} &= \frac{1-\half a\Delta t}{1 + \half a\Delta t} u^n & \hbox{Crank-Nicolson}
\label{decay:CN:u}\\ 
u^{n+1} &= \frac{1 - (1-\theta) a\Delta t}{1 + \theta a \Delta t}u^n  & \hbox{The }\theta-\hbox{rule}
\label{decay:th:u}
\end{alignat}
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Test the understanding!}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Compact operator notation for finite differences}

\label{decay:fd:op}

@@

\noindent
\begin{equation}
[D_t^- u  = -au]^n
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Specific notation for difference operators}

@@

Backward difference:
\begin{equation}
[D_t^-u]^n = \frac{u^{n} - u^{n-1}}{\Delta t}
\approx \frac{d}{dt} u(t_n) \label{fd:D:b}
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Backward Euler scheme with operator notation}

@@

\begin{equation}
[D_t^- u  = -au]^n
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Forward Euler scheme with operator notation}

\begin{equation}
[D_t^+ u  = -au]^n
\end{equation}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The Crank-Nicolson scheme with operator notation}

@@

Test: use the definitions and write out the above formula to see that
it really is the Crank-Nicolson scheme!
\end{frame}

\section{Implementation}

\begin{frame}[plain,fragile]
\frametitle{Implementation}

\label{decay:impl1}

@@

\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Requirements of a program}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Tools to learn}

@@

\noindent
\begin{block}{Notice}
All programs are in the directory
\href{{http://tinyurl.com/ofkw6kc/alg}}{\nolinkurl{src/alg}}.
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Why implement in Python?}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Why implement in Python?}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Algorithm}

\label{decay:py1}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Translation to Python function}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *

>>>>

%%
%% Automatically generated file from DocOnce source
%% (https://github.com/hplgit/doconce/)
%%

% #define PREAMBLE

% #ifdef PREAMBLE
%-------------------- begin preamble ----------------------

\documentclass[%
oneside,                 % oneside: electronic viewing, twoside: printing
final,                   % draft: marks overfull hboxes, figures with paths
10pt]{article}

\listfiles               % print all files needed to compile this document

\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
\usepackage[table]{xcolor}
\usepackage{bm,microtype}

\usepackage[pdftex]{graphicx}

% Packages for typesetting blocks of computer code
\usepackage{fancyvrb,framed,moreverb}

% Define colors
\definecolor{orange}{cmyk}{0,0.4,0.8,0.2}
\definecolor{tucorange}{rgb}{1.0,0.64,0}
\definecolor{darkorange}{rgb}{.71,0.21,0.01}
\definecolor{darkgreen}{rgb}{.12,.54,.11}
\definecolor{myteal}{rgb}{.26, .44, .56}
\definecolor{gray}{gray}{0.45}
\definecolor{mediumgray}{gray}{.8}
\definecolor{lightgray}{gray}{.95}
\definecolor{brown}{rgb}{0.54,0.27,0.07}
\definecolor{purple}{rgb}{0.5,0.0,0.5}
\definecolor{darkgray}{gray}{0.25}
\definecolor{darkblue}{rgb}{0,0.08,0.45}
\definecolor{darkblue2}{rgb}{0,0,0.8}
\definecolor{lightred}{rgb}{1.0,0.39,0.28}
\definecolor{lightgreen}{rgb}{0.48,0.99,0.0}
\definecolor{lightblue}{rgb}{0.53,0.81,0.92}
\definecolor{lightblue2}{rgb}{0.3,0.3,1.0}
\definecolor{lightpurple}{rgb}{0.87,0.63,0.87}
\definecolor{lightcyan}{rgb}{0.5,1.0,0.83}

\colorlet{comment_green}{green!50!black}
\colorlet{string_red}{red!60!black}
\colorlet{keyword_pink}{magenta!70!black}
\colorlet{indendifier_green}{green!70!white}

% Backgrounds for code
\definecolor{cbg_gray}{rgb}{.95, .95, .95}
\definecolor{bar_gray}{rgb}{.92, .92, .92}

\definecolor{cbg_yellowgray}{rgb}{.95, .95, .85}
\definecolor{bar_yellowgray}{rgb}{.95, .95, .65}

\colorlet{cbg_yellow2}{yellow!10}
\colorlet{bar_yellow2}{yellow!20}

\definecolor{cbg_yellow1}{rgb}{.98, .98, 0.8}
\definecolor{bar_yellow1}{rgb}{.98, .98, 0.4}

\definecolor{cbg_red1}{rgb}{1, 0.85, 0.85}
\definecolor{bar_red1}{rgb}{1, 0.75, 0.85}

\definecolor{cbg_blue1}{rgb}{0.87843, 0.95686, 1.0}
\definecolor{bar_blue1}{rgb}{0.7,     0.95686, 1}

%\setlength{\fboxsep}{-1.5mm}  % adjust cod_vpad/pro_vpad background box

%% Background for code blocks (parameter is color name)

%% pro/cod_vpad: gives some vertical padding before and after the text
%% (but has more simplistic code than _cod/pro_tight+cod/pro).
%% pro/cod_vpad can be used to enclose Verbatim or lst begin/end for code.
%% pro/cod calls _pro/cod_tight and has very little vertical padding,
%% used to enclose Verbatim and other begin/end for code.
%% (pro/cod is what the ptex2tex program could produce with the
%% Blue/BlueBar definitions in .ptex2tex.cfg.)

\newenvironment{cod_vpad}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_cod_tight}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{cod}[1]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_cod_tight}{#1}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_cod_tight}\egroup}

%% Background for complete program blocks (parameter 1 is color name
%% for background, parameter 2 is color for left bar)
\newenvironment{pro_vpad}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_pro_tight}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\advance\hsize-2mm\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{pro}[2]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_pro_tight}{#1}{#2}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_pro_tight}\egroup}

\usepackage{listingsutf8}

% Common lstlisting parameters
\lstset{
  basicstyle=\small \ttfamily,
  breaklines=false,          % break/wrap lines
  breakatwhitespace=true,    % let linebreaks happen at whitespace
  breakindent=40pt,
  tab=,
  tabsize=4,                 % tab means 4 spaces
  %belowskip=\smallskipamount,  % space between code and text below
  xleftmargin=5pt,           % indentation of code frame
  xrightmargin=5pt,
  framexleftmargin=5pt,      % add frame space to the left of code
  %numbers=left,             % put line numbers on the left
  %stepnumber=2,             % stepnumber=1 numbers each line, =n every n lines
  %framerule=0.4pt           % thickness of frame
  aboveskip=2ex,             % vertical space above code frame
  showstringspaces=false,    % show spaces in strings with an underscore
  showspaces=false,          % show spaces with an underscore
  showtabs=false,
  keepspaces=true,
  columns=fullflexible,      % tighter character kerning, like verb
  escapeinside={||},         % for |\pause| in slides and math in code blocks
  extendedchars=\true,       % allows non-ascii chars, does not work with utf-8
}

% Internally defined styles for lstlisting

\lstdefinestyle{simple}{
commentstyle={},
}

% end of custom lstdefinestyles

\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern

% Hyperlinks in PDF:
\definecolor{linkcolor}{rgb}{0,0,0.4}
\usepackage{hyperref}
\hypersetup{
    breaklinks=true,
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    citecolor=black,
    filecolor=black,
    %filecolor=blue,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
    }
%\hyperbaseurl{}   % hyperlinks are relative to this root

\setcounter{tocdepth}{2}  % number chapter, section, subsection

% Tricks for having figures close to where they are defined:
% 1. define less restrictive rules for where to put figures
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{0.75}
% floatpagefraction must always be less than topfraction!
% 2. ensure all figures are flushed before next section
\usepackage[section]{placeins}
% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
%\usepackage{float}\restylefloat{figure}

\usepackage[framemethod=TikZ]{mdframed}

% --- begin definitions of admonition environments ---

% --- end of definitions of admonition environments ---

% prevent orhpans and widows
\clubpenalty = 10000
\widowpenalty = 10000

\newenvironment{doconceexercise}{}{}
\newcounter{doconceexercisecounter}

% ------ header in subexercises ------
%\newcommand{\subex}[1]{\paragraph{#1}}
%\newcommand{\subex}[1]{\par\vspace{1.7mm}\noindent{\bf #1}\ \ }
\makeatletter
% 1.5ex is the spacing above the header, 0.5em the spacing after subex title
\newcommand\subex{\@startsection*{paragraph}{4}{\z@}%
                  {1.5ex\@plus1ex \@minus.2ex}%
                  {-0.5em}%
                  {\normalfont\normalsize\bfseries}}
\makeatother

% --- end of standard preamble for documents ---

@@

\raggedbottom
\makeindex
\usepackage[totoc]{idxlayout}   % for index in the toc
\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc

@@

\thispagestyle{empty}

\begin{center}
{\LARGE\bf
\begin{spacing}{1.25}
Study guide: Algorithms and implementations for exponential decay models
\end{spacing}
}
\end{center}

@@

\begin{center}
{\bf Hans Petter Langtangen${}^{1, 2}$} \\ [0mm]
\end{center}

\begin{center}
% List of all institutions:
\centerline{{\small ${}^1$Center for Biomedical Computing, Simula Research Laboratory}}
\centerline{{\small ${}^2$Department of Informatics, University of Oslo}}
\end{center}

% ----------------- end author(s) -------------------------

% --- begin date ---
\begin{center}
Oct 10, 2015
\end{center}
% --- end date ---

\vspace{1cm}

% !split
\section*{INF5620 in a nutshell}
\label{5620:about}

@@

% --- begin paragraph admon ---
\paragraph{After the course.}
You see a PDE and can't wait to program a method
and visualize a solution! Somebody asks if the solution is right
and you can give a convincing answer.
% --- end paragraph admon ---

% !split
\subsection*{The new official six-point course description}

@@

% !bpop
\begin{itemize}
 \item can derive methods and implement them to solve frequently
   arising partial differential equations (PDEs) from physics and mechanics.

 \item have a good understanding of finite difference and finite element
   methods and how they are applied in linear and nonlinear PDE problems.

 \item can identify numerical artifacts and perform mathematical analysis
   to understand and cure non-physical effects.

 \item can apply sophisticated programming techniques in Python, combined
   with Cython, C, C++, and Fortran code, to create modern,
   flexible simulation programs.

 \item can construct verification tests and automate them.

 \item have experience with project hosting sites (GitHub),
   version control systems (Git), report writing ({\LaTeX}),
   and Python scripting for performing reproducible computational science.
\end{itemize}

\noindent
% !epop

% !split
\subsection*{More specific contents: finite difference methods}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
   \item ODEs

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{More specific contents: finite element methods}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
   \item stationary diffusion equations $u_{xx}=f$ in 1D

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{More specific contents: nonlinear and advanced problems}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
 \item Nonlinear PDEs
\begin{itemize}

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{Philosophy: simplify, understand, generalize}

@@

\noindent
This is the power of applied mathematics!

% !split
\subsection*{The exam}

% !bpop
\begin{itemize}
 \item Oral exam

 \item 6 problems (topics) are announced two weeks before the exam

 \item Work out a 20 min presentations (talks) for each problem

 \item At the exam: throw a die to pick your problem to be presented

 \item Aids: plots, computer programs

 \item Why? Very effective way of learning

 \item Sure? Excellent results over 15 years

 \item When? Late december
\end{itemize}

\noindent
% !epop

% !split
\subsection*{Required software}

@@

\noindent
% !split
\subsection*{Assumed/ideal background}

@@

\noindent
% !split
\subsection*{Start-up example for the course}

@@

\noindent
% !split
\subsection*{Start-up example}

% --- begin paragraph admon ---
\paragraph{ODE problem.}
\[ u'=-au,\quad u(0)=I,\ t\in (0,T],\]
where $a>0$ is a constant.
% --- end paragraph admon ---

Everything we do is motivated by what we need as building blocks for
solving PDEs!

% !split
\subsection*{What to learn in the start-up example; standard topics}

@@

\noindent
% !split
\subsection*{What to learn in the start-up example; programming topics}

@@

\noindent
% !split
\subsection*{What to learn in the start-up example; mathematical analysis}

@@

\noindent
% !split
\subsection*{What to learn in the start-up example; generalizations}

@@

\noindent
% !split
\section*{Finite difference methods}
\label{decay:fdm}

@@

% !split
\subsection*{Topics in the first intro to the finite difference method}

% --- begin paragraph admon ---
\paragraph{Contents.}
\begin{itemize}
 \item How to think about finite difference discretization

@@

\noindent
% --- end paragraph admon ---

% !split
\subsection*{A basic model for exponential decay}
\label{decay:model}

@@

% --- begin paragraph admon ---
\paragraph{Observation.}
We can learn a lot about numerical methods, computer implementation,
program testing, and real applications of these tools by using
this very simple ODE as example. The teaching principle is to keep the math as
simple as possible while learning computer tools.
% --- end paragraph admon ---

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/FE1.png}}

% !split
\subsection*{The ODE model has a range of applications in many fields}

@@

\noindent
See the \href{{http://tinyurl.com/nclmcng/pub/sphinx-decay/._main_decay008.html}}{text} for details.

% !split
\subsection*{The ODE problem has a continuous and discrete version}

\paragraph{Continuous problem.}
\begin{equation}
u' = -au,\ t\in (0,T], \quad u(0)=I
\label{decay:problem}
\end{equation}
(varies with a continuous $t$)

\paragraph{Discrete problem.}
Numerical methods applied to the continuous problem turns it into
a discrete problem

\begin{equation}
u^{n+1} = \mbox{const} u^n, \quad n=0,1,\ldots N_t-1, \quad u^n=I
\label{decay:problem:discrete}
\end{equation}
(varies with discrete mesh points $t_n$)

% !split
\subsection*{The steps in the finite difference method}
\label{decay:schemes:keysteps}

@@

\noindent
% !split
\subsection*{Step 1: Discretizing the domain}

@@

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/fdm_u_ue.png}}

% !split
\subsection*{What about a mesh function between the mesh points?}

@@

% inline figure
\centerline{\includegraphics[width=0.7\linewidth]{fig-alg/fdm_u_uei.png}}

% !split
\subsection*{Step 2: Fulfilling the equation at discrete time points}

@@

\index{finite differences}

% !split
\subsection*{Step 3: Replacing derivatives by finite differences}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_forward.pdf}}

% !split
\subsection*{Step 3: Replacing derivatives by finite differences}

@@

(Known as \emph{discrete equation}, or \emph{discrete problem},
or \emph{finite difference method/scheme})

% !split
\subsection*{Step 4: Formulating a recursive algorithm}

@@

% --- begin paragraph admon ---
\paragraph{The Forward Euler scheme.}
Solve wrt $u^{n+1}$ to get the computational formula:

\begin{equation}
u^{n+1} = u^n - a(t_{n+1} -t_n)u^n
\label{decay:FE}
\end{equation}
% --- end paragraph admon ---

% !split
\subsection*{Let us apply the scheme by hand}

@@

Ooops - we can find the numerical solution by hand (in this simple
example)! No need for a computer (yet)...

% !split
\subsection*{A backward difference}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_backward.pdf}}

% !split
\subsection*{The Backward Euler scheme}

@@

\begin{equation}
u^{n+1} = \frac{1}{1+ a(t_{n+1}-t_n)} u^n
\label{decay:BE}
\end{equation}

% !split
\subsection*{A centered difference}

@@

% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-alg/fd_centered_CN.pdf}}

% !split
\subsection*{The Crank-Nicolson scheme; ideas}
\label{decay:schemes:CN}

@@

\[ u(t_{n+\half}) \approx \half(u^n + u^{n+1}) \]

% !split
\subsection*{The Crank-Nicolson scheme; result}

@@

\begin{equation}
u^{n+1} = \frac{1-\half a(t_{n+1}-t_n)}{1 + \half a(t_{n+1}-t_n)}u^n
\label{decay:CN}
\end{equation}
This is a Crank-Nicolson (CN) scheme or a midpoint or centered scheme.

% !split
\subsection*{The unifying $\theta$-rule}
\label{decay:schemes:theta}

@@

\begin{equation}
u^{n+1} = \frac{1 - (1-\theta) a(t_{n+1}-t_n)}{1 + \theta a(t_{n+1}-t_n)}
\label{decay:th}
\end{equation}

% !split
\subsection*{Constant time step}

@@

% --- begin paragraph admon ---
\paragraph{Summary of schemes for constant time step.}
\begin{alignat}{2}
u^{n+1} &= (1 - a\Delta t )u^n  & \hbox{Forward Euler}
\label{decay:FE:u}\\ 
u^{n+1} &= \frac{1}{1+ a\Delta t} u^n  & \hbox{Backward Euler}
\label{decay:BE:u}\\ 
u^{n+1} &= \frac{1-\half a\Delta t}{1 + \half a\Delta t} u^n & \hbox{Crank-Nicolson}
\label{decay:CN:u}\\ 
u^{n+1} &= \frac{1 - (1-\theta) a\Delta t}{1 + \theta a \Delta t}u^n  & \hbox{The }\theta-\hbox{rule}
\label{decay:th:u}
\end{alignat}
% --- end paragraph admon ---

% !split
\subsection*{Test the understanding!}

@@

\noindent
% !split
\subsection*{Compact operator notation for finite differences}
\label{decay:fd:op}

@@

\noindent
\begin{equation}
[D_t^- u  = -au]^n
\end{equation}

% !split
\subsection*{Specific notation for difference operators}

@@

Backward difference:
\begin{equation}
[D_t^-u]^n = \frac{u^{n} - u^{n-1}}{\Delta t}
\approx \frac{d}{dt} u(t_n) \label{fd:D:b}
\end{equation}

% !split
\subsection*{The Backward Euler scheme with operator notation}

@@

\begin{equation}
[D_t^- u  = -au]^n
\end{equation}

% !split
\subsection*{The Forward Euler scheme with operator notation}

\begin{equation}
[D_t^+ u  = -au]^n
\end{equation}

% !split
\subsection*{The Crank-Nicolson scheme with operator notation}

@@

Test: use the definitions and write out the above formula to see that
it really is the Crank-Nicolson scheme!

% !split
\section*{Implementation}
\label{decay:impl1}

@@

\noindent
% !split
\subsection*{Requirements of a program}

@@

\noindent
% !split
\subsection*{Tools to learn}

@@

% --- begin paragraph admon ---
\paragraph{Notice.}
All programs are in the directory
\href{{http://tinyurl.com/ofkw6kc/alg}}{\nolinkurl{src/alg}}.
% --- end paragraph admon ---

% !split
\subsection*{Why implement in Python?}

@@

\noindent
% !split
\subsection*{Why implement in Python?}

@@

\noindent
% !split
\subsection*{Algorithm}
\label{decay:py1}

@@

\noindent
% !split
\subsection*{Translation to Python function}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *
